Adaptive control of action timing with a spiking neural integrator model of anterior cingulate cortex

### 2013-06-10 paper/jneurosci2013.tex
Third version of the paper, with figures.

%\documentclass[twocolumn]{article}
\documentclass[12pt]{article}
% Change "article" to "report" to get rid of page number on title page
\usepackage{amsmath,amsfonts,amsthm,amssymb}
\usepackage{setspace}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{extramarks}
\usepackage{chngpage}
\usepackage{soul}
\usepackage[usenames,dvipsnames]{color}
\usepackage{graphicx,float,wrapfig}
\usepackage{listings}
\usepackage{framed}
\usepackage{inconsolata}
\usepackage{cite}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Make title
\title{}
\date{}
\author{}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

\begin{center} \begin{Large}
  \textbf{A spiking neural circuit model of cortical function
          in simple reaction time tasks}
\end{Large} \end{center}

\begin{center}
  Trevor Bekolay, Benjamine Liu, Nandakumar Narayanan, Chris
  Eliasmith, Mark Laubach
\end{center}

\begin{abstract}
  todo
\end{abstract}

@@

Animals make complicated motor actions
with high spatial and temporal resolutions.
In temporally precise tasks,
the motor cortex and several afferents coordinate
to bring about motor actions at appropriate times.
For tasks with delay periods,
the medial prefrontal cortex is critically important
for suppressing motor action during the delay period.
This effect has been seen in ?.  %???primates humans rodents etc???.

In rodents, the role of the medial prefrontal cortex
has been investigated through ?.
%???previous work???.
Previously, we have used a simple reaction task
and principal component analysis
to investigate ?. %???summarize Narayana \& laubach jnp 2009???
These results point to neural integration
as playing a key role
in the mPFC's ability
to control the timing of motor actions.

In order to investigate the neural circuit
underlying the production and suppression of motor output,
we extend Singh \& Eliasmith's model to include
neural populations that directly affect the environment,
analogous to the function of Betz cells.
In doing so, we can relate the model to the experimental data
in both neural and behavioural contexts.

\section{Model}

The model presented in this study
is based on a conceptual model of PFC and MC
function in delay tasks proposed by Narayanan and Laubach.
Their conceptual model is based on
principal component analysis of neural data
in rats performing a simple reaction time task.

Figure 1 summarizes their PC analysis.
In PFC, two components account for over 35\%
of variance in population firing.
The first principal component increases slightly
at the time of the lever press,
and then decreases over the delay period.
The second principal component
also increases at the time of lever press,
but remains high until the time of release.

In motor cortex, three components account
for over 60\% of variance in population firing.
The first two principal components are
similar to those found in PFC;
the first steadily decreases over the delay period,
and the second remains high during the delay period.
The third principal component increases
sharply slightly before lever press
and lever release.

The identification of these principal components
leads to a conceptual model of cortical function
during simple reaction time tasks.
This model is summarized in figure 2. % Fig
At the start of a trial,
two complementary systems are activated
to produce quick reactions at the time of a stimulus.
The \textit{prepotent response} system
prepares the motor action that the subject will perform
at the end of the delay period.
The \textit{proactive inhibition} system
suppresses that motor action until the delay period
is over.

\subsection{Neural integration as a core mechanism}

An important observation of Narayanan and Laubach
is that the second principal component
is the cumulative sum (integral)
of the first principal component.
This suggests that some neurons in these cortical areas
are representing the integral of another
signal represented in these areas.

Neural integration through recurrently connected
populations of neurons has been widely studied
in the context of working memory.
% ? cite a bunch -- Mawkinson, Roman, Tanji, X-J Wang, look at singh eliasmith biblio

The Singh and Eliasmith model, in particular,
facilitates a neural implementation of the conceptual model
proposed as a result of the PC analysis.
In addition to outlining a method of constructing
multidimensional neural integrators,
it also provides a method of connecting those integrators
to other populations of neuron representing other signals.

\subsection{Coupled integrators in a control system}

With non-integrative and integrative neuronal populations
as the building blocks, we propose a neural circuit
realization of the conceptual model
discussed in Narayanan and Laubach.
The model is summarized in figure 3.

Two populations are hypothesized to reside in motor cortex.
The \textit{Press/Release} population represents
whether the subject is currently executing
a press or release motor action.
The activity of this population directly affects the environment
in the model, which can be thought of
as functioning like Betz cells.
The \textit{Holding} population tracks
whether or not the subject is currently holding down the lever.
It is analogous to the \textit{prepotent response}
from the conceptual model, as in the absence of inhibition,
it will activate the \textit{Release} neurons
in the \textit{Press/Release} population.
The \textit{Holding} population is integrative,
and integrates the
\textit{Press} signal from the \textit{Press/Release}
population. The integration is controlled
by the \textit{Release} signal, meaning that
when the lever is released the \textit{Holding} signal
is reset to some baseline.

Three populations are hypothesized to reside
in medial prefrontal cortex.
The \textit{Waiting} population represents
that the subject is in a delay state,
waiting for the appropriate time to perform a motor action.
It is analogous to the \textit{proactive inhibition}
from the conceptual model, as it works to balance
the prepotent release response being signaled
from the motor cortex's \textit{Holding} population.
The \textit{Cue} population is sensitive
to the auditory stimulus delivered to the subject
when the delay period is over.
It inhibits the \textit{Waiting} population's
proactive inhibition, disinhibiting \textit{Release}
sensitive neurons.
Similarly, the \textit{Timer} population
tracks the relative amount of time that has passed
since the subject pressed down the lever,
starting the trial.
When the time of the stimulus can be predicted,
the \textit{Timer} population speeds
responding by increasingly inhibiting the
\textit{Waiting} population over time.
It can also effect a motor action on its own by
disinhibiting \textit{Release} neurons sufficiently.
The \textit{Timer} population is integrative,
gradually ramping up as it integrates
the \textit{Waiting} signal.

@@

In both the empirical and simulated data,
subjects and models performed the same simple reaction time task.
At the start of a trial,
subjects press down a lever.
A timer tracks the time since the lever was pressed down,
and after an experimenter-specified time period,
called the \textit{delay period} or \textit{foreperiod},
an auditory cue signals that the subject must respond
by releasing the lever.
If the lever is released within 600 milliseconds,
the trial is considered a success
and reward is delivered.
If the lever is released before the auditory cue
or after the 600 millisecond response window,
then the trials is considered a failure
and subjects are given a time out in which
lights are extinguished.
Lever presses during this timeout period
lengthen the timeout period.

Empirically, this study is done with rats
in a MedPC box containing a lever,
house lights, and a spout for delivering water reward.
Rats were trained by first
associating reward with lever presses,
and then introducing progressively longer
delay periods until they reached the goal delay
of 1 second.
They were considered to have learned the task
when they performed the task correctly $>60\%$
of the time.

Theoretically, this study was performed
in the Nengo simulation environment.
Subjects were networks of leaky integrate-and-fire
neurons, as described below.
The environment was modeled using a
\texttt{simplenode}, which allow arbitrary code
to interact with a neural model.
In this case, the \texttt{simplenode} implemented
a finite state machine to track task state,
and scalar values to indicate the state of
house lights, lever position, auditory cue, and reward.

\subsection{Empirical summary}

The empirical study (described in more depth in ?) %???cite???)
was done with twelve male Long-Evans rats (3-4 months old; Harlan).
Rats were water restricted, with food always available,
and consumed 10-15 ml of water during each
behavioural session.

Arrays of recording electrodes were implanted
in dmPFC ($n = 8$), motor cortex ($n = 3$), or both ($n = 1$).
Electrodes were 50 $\mu$m stainless steel wires
with 250 $\mu$m between wires.
Arrays were in configurations of
4 $\times$ 4 or 3 $\times$ 3 $\times$ 2.
Microwire arrays were targeted at dmPFC
based on the following coordinates from bregma:
anteroposterior [AP]: $+$3.2mm;
mediolateral [ML]: $\pm$1.4mm;
dorsoventral [DV]: $-$3.6mm at 10$^\circ$ in the frontal plane.
Motor cortex was targeted with
AP: $-$0.5mm; ML: $\pm$2.5-3.5mm;
DV: $-$1.5mm at $-25^\circ$ in the frontal plane.
Histology was performed to confirm electrode locations
(???see fig???).

Ensemble recordings were done using a
Many Neuron Acquisition Program (Plexon, Dallas, TX).
Spike sorting using principal component analysis (PCA)
and waveform shape identified 414 single units
whose spikes were included in this study.

\subsection{Theoretical}

Two models were created to explain neural processing
in the simple reaction time task.
The first is a double-integrator model,
as is described in ?, %???cite singh eliasmith???,
which allows for precise comparisons
between the spiking behaviour of
the model and the empirical subjects
given a controllable behavioural outcome.
The second is an integrated model that
expands the double-integrator model
to include populations that manipulate the environment,
allowing for the comparison between
the model and empirical behaviour.

Both models are made up of heterogeneous populations
of leaky integrate-and-fire neurons
created and connected using the methods
of Eliasmith and Anderson ?. % ???cite???.
Connections between populations of neurons
are determined based on the function computed
through the connection.
Connection weights are determined
by solving a least-squares minimization
of the function representation
given the tuning properties
of the neurons representing the function parameters,
as described in ? %???singheliasmith???
and in detail in ?. %???eliasmith anderson ???.

@@

In  %???cite???
it is shown that
a population of heterogeneous adaptive leaky integrate-and-fire neurons
representing a two-dimensional signal
can capture the dynamics of
working memory cells in prefrontal cortex.
There are, however, several network configurations
that can result in a population of neurons
representing two integrated signals,
as shown in figure ?. %???copy fig 5 from paper???.

We adopt the coupled one-dimensional integrator model,
despite little evidence that reciprocal projections
are localized to smaller integrative populations.
Instead, it is the PCA of mPFC activity
that suggests many neurons are
only sensitive to one of the principal components.
Additionally, it is more intuitive
to connect each principal component with
a separate neural population.
The important task, then,
is to connect each population
with the appropriate value being encoded.

>>>>

\documentclass[11pt]{article}
% Change "article" to "report" to get rid of page number on title page
\usepackage{amsmath,amsfonts,amsthm,amssymb}
\usepackage{setspace}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{extramarks}
\usepackage{chngpage}
\usepackage{soul}
\usepackage[round]{natbib}
\usepackage[usenames,dvipsnames]{color}
\usepackage{graphicx,float,wrapfig}
\usepackage{listings}
\usepackage{framed}
\usepackage{inconsolata}
\usepackage{hyperref}

\hypersetup{
  colorlinks = true, %Colours links instead of ugly boxes
  urlcolor = blue, %Colour for external hyperlinks
  linkcolor = blue, %Colour of internal links
  citecolor = red %Colour of citations
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Make title
\title{Adaptive control of action timing with a spiking neural integrator
  model of anterior cingulate cortex}
\date{}
\author{Trevor Bekolay, Chris Eliasmith, Mark Laubach}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

\maketitle

\begin{abstract}
  Animals can learn to perform a simple reaction time task
  with a predictable foreperiod
  by predicting the time of the cue,
  achieving reaction times faster
  than through cue-responding alone.
  The outcome of the previous trial
  is one factor in determining
  whether an animal will attempt to time
  the foreperiod.
  The anterior cingulate cortex (ACC) has been implicated
  in enabling control of action timing
  depending on the outcome of the previous trial.
  Analysis of ACC activity suggests that
  neural integration is a key mechanism
  for adaptive control in precisely timed tasks.
  We show that a spiking neural circuit
  consisting of coupled neural integrators
  captures the neural dynamics
  of the experimentally recorded ACC.
  We use this coupled integrator model
  to adaptively control
  a spiking neural circuit that follows
  a cue-response strategy,
  and show that it performs the simple reaction time task
  with similar reaction times as experimental subjects.
  We then alter the model
  in a way consistent with
  how the aged brain degrades,
  and predict that aged subjects
  will adopt a cue-response strategy.
\end{abstract}

@@

The anterior cingulate cortex has been implicated
in a wide variety of functions,
including conflict monitoring
\citep{Aarts2009, Botvinick2004, VanVeen2004},
error monitoring \citep{Brown2005, Carter1998, Fecteau2003,
  Modirrousta2008, Rushworth2004},
anticipatory control \citep{Aarts2008, Koyama2001},
arousal of the sympathetic nervous system
\citep{Critchley2003, Luu2003},
and control of action timing \citep{Muir1996, Mulert2003,
  Naito2013, Narayanan2009, Risterucci2003}.
Many have proposed integrative or general models
that would allow the ACC to be involved
in several of these functions,
depending on the behavioral context
\citep{Botvinick2004, Luu2003}.
However, while there have been
attempts at modeling the ACC at the level
of artificial neural networks
\citep{Botvinick2004, Brown2005},
there are currently no mathematical
accounts of ACC function that would
explain the myriad of behaviors
modulated by the ACC.
Similarly, no existing models
have been implemented with spiking neurons,
and are therefore not directly comparable
to neural activity.

In this paper, we propose that a
two-dimensional dynamical system
can capture the activity of the ACC.
We test the model in a simple reaction time task
with a predictable foreperiod and
show that the mathematical model
enables precise timing of actions,
and changes behavior depending
on the outcome of the last trial.
We implement the dynamical system
in a spiking neural network
based on \citet{Singh2006}
and show that simulated neural activity
closely matches recorded ACC activity.
We show that the model
can be combined with
a motor control model,
resulting in behavioral changes
analogous to previous behavioral theories
of ACC function.
Finally, we damage the model in a way analogous
to aging, and predict that
aged subjects will adopt a more conservative
strategy due to a reduction of the
ability of the ACC to control action timing.

Our mathematical account of ACC function
is applied directly to
error monitoring and
the control of action timing.
However, we assert that the dynamical systems
approach is general and can be used
to explain other theories
of high-level ACC function.

\section{Simple reaction time task} \label{sec:simple-rt}

The simple reaction time task
serves as a minimal task in which
to examine the control of precisely timed actions.
In this task, the subject presses and holds down a lever
for a fixed amount of time (the ``foreperiod'';
1 second for this study).
After the foreperiod, a cue is presented
(e.g., an auditory tone).
The subject then has 0.6 seconds (the response window)
in which to release the lever.
If the lever is released in that time,
the trial is a success, and the subject is rewarded.
If the lever is released during the foreperiod,
the trial is classified as a ``premature error,''
and is penalized by a time-out.
If the lever is not released within the response window,
the trail is classified as a ``late error,''
and is also penalized by a time-out.
Figure~\ref{fig1} presents a schematic view of the task.

\section{Analysis of experimental data}

\citet{Narayanan2009} recorded
the anterior cingulate cortex (ACC) and motor cortex
of rodents performing this simple reaction-time task.
They used principal component analysis
on normalized spike densities
in order to determine
what is being represented in these brain areas
during this task.
As shown in Figure~\ref{fig2},
the first two principal components
found from analyzing the ACC data are large,
and are hypothesized to track
1) if the subject was performing the task, and
2) the relative amount of time the subject has been
performing the task.
\citeauthor{Narayanan2009} also found that the neural activity
in the ACC changed significantly depending on
the outcome of the previous trial;
the first two principal components
change significantly after errors,
and appear to contain information that an error occurred.
This information would be necessary
in order to adapt behavior based on the last trial.

Interestingly, in the post-correct case,
the first principal component
is highly correlated with the integral
of the second principal component,
and vice versa.
In the post-error case,
these correlations are significantly weaker.
This result suggests that
the ACC is performing some kind of integration..

\section{Double integrator model}

One model of neural integration
was proposed in the context of working memory
by \citet{Singh2006}.
Their model, called the \textit{double integrator} model,
uses multidimensional neurons
in integrative populations
to exhibit time-varying activity
during a vibrotactile discrimination task.
It relies on multiple coupled integrators,
whose activity resembles
the two strong principal components
in the ACC data.
The model employs
the Neural Engineering Framework (NEF; \citealp{Eliasmith2003})
to implement a spiking neural network
that captures the dynamics of the
integrative dynamical system
(see section \ref{sec:nef}).
We hypothesize that the \citeauthor{Singh2006} model
will exhibit the same neural dynamics
as the experimental ACC in the simple RT task.

\subsection{Dynamics}

The dynamics of the double integrator model
are captured in the following dynamics equations.
\begin{align*}
  \dot{x}_1 &= u \\
  \dot{x}_2 &= \alpha x_1,
\end{align*}
where $u$ represents an input signal
and $\alpha$ is the strength of
the connection between the two integrators.
This results in a two-dimensional system
in which $x_1$ integrates its input,
maintaining the accumulated value over time.
$x_2$ integrates the value of $x_1$.
\citet{Singh2006} have proposed that this system
enables time tracking;
because $x_2$ integrates $x_1$,
the amount of time elapsed since
$x_1$ changes can be determined
based on how much $x_2$ has changed.

In the simple reaction time task,
the input, $u$, is the activation of neurons
that effect the lever press.
The system is also modified by signals
representing the outcome of a particular trial;
on a correct trial, reward delivery ($R > 0$)
resets the system to a starting state.
On an error trail, a second input signal ($E < 0$)
drives $x_1$ to a low state.
When implemented in a spiking network,
this causes $x_2$ to saturate at a low value,
effectively breaking the integration,
as is seen in Figure~\ref{fig3} in post-error trails.
This results in the following dynamics.
\begin{align} \label{dint}
  \dot{x}_1 &= u + E - R x_1 \nonumber \\
  \dot{x}_2 &= \alpha x_1 - R x_2.
\end{align}
See Figure~\ref{fig4} for a graphical
depiction of the double integrator model.

In addition to the depicted system,
we inject a slowly oscillating signal into $x_1$.
This is motivated by the observation
that system \eqref{dint} is stationary
during the intertrial interval;
$x_1$ is either near 0 on rewarded trials
or some low value on error trials.
$x_2$ does not change on rewarded trials,
and saturates at some low value on error trials.
Figure~\ref{fig3} shows some activity during
the intertrial interval;
we model this additional activity
by a simple oscillation, $O = \sin(2 \pi f t + \phi)$,
where the frequency $f$ is low ($\le 0.4$Hz).
The final dynamics are
\begin{align} \label{dint-full}
  \dot{x}_1 &= u + E + O - R x_1 \nonumber \\
  \dot{x}_2 &= \alpha x_1 - R x_2.
\end{align}

Note that while there is experimental evidence for
a slow oscillations like $O$ in various brain regions
\citep{Buzsaki2004}, this injected oscillation
does not affect the performance of the network,
and is intended to model ACC activity
not being otherwise modelled in the current study.
This activity could reflect
other functions that the ACC is performing,
or could reflect the overall pace
of the simple RT task.

\section{Adaptive control circuit}

\citet{Narayanan2009} hypothesized
that the ACC activity analyzed in the present study
could be used to inhibit motor responses
until the time of the cue.
In order to verify that hypothesis,
we constructed a network that
performs the simple RT task
using the simple non-adaptive strategy
of responding to cues.
We then combined that network
with the double integrator network
such that the final model
is able to flexibly switch between
an aggressive timing strategy
and a conservative cue-response strategy
depending on the outcome of the previous trial.

\subsection{Cue-responding circuit}

The cue-responding circuit is composed
of three populations
of adaptive leaky integrate-and-fire neurons
(see Figure~\ref{fig4}).
The population labelled \textit{Trigger}
is sensitive to the cue.
The population labelled \textit{Holding}
tracks whether the lever is currently being held down.
The population labelled \textit{Press/Release}
causes lever presses and releases
through activation of different subsets
of its neurons.
The actual lever press and release
is delayed by $\sim$250ms,
the approximate amount of time
it takes for muscles to respond
to the activation of command neurons
in primary motor cortex.  % ???travis

The \textit{Holding} neurons project
to the \textit{Trigger} population
to provide the behavioral context
(i.e., \textit{can} I release the lever?)
The \textit{Trigger} neurons send projections
to the \textit{Press/Release} neurons
upon a cue that occurs in the correct context
(i.e., \textit{should} I release the lever?)

The context-driven design of this circuit
provides an alternative theory to
the prepotent inhibition model
proposed by \citet{Narayanan2009}.
They suggest that a motor command
is activated at the same time
as an inhibitory waiting signal,
and it is the removal of that inhibition
that produces the motor command.
This model proposes that
an excitatory context signal is activated,
and the pairing of that context signal
with the cue produces the motor command.
This leads to an alternative hypothesis
for the function of ACC activity:
the ACC tracks the behavioral context,
and is able to inhibit or excite motor cortex,
but does not always necessarily do so.

\subsection{Adaptive control circuit}

A weakness of the contextual cue-responding circuit
is that there is a fixed lower-bound
on the reaction time
based on the time taken to detect the cue
and for the delay between activation
of motor cortex and muscle activity.
Since the goal is to time the action
as close as possible to the time of the cue,
the cue-responding strategy may be too slow,
and so a predictive approach must be employed.

The double integrator model described
previously can implement such a strategy.
The state of the dynamical system being tracked by
the double integrator (equation \eqref{dint-full})
is approximately the same when the cue occurs.
By connecting the output of the double integrator
to the \textit{Trigger} population,
the double integrator can cause a lever release
when the state of the double integrator
is slightly before the predicted
time of the cue.

However, if the previous trial was an error,
the double integrator will not
be following the trajectory
that predicts the time of the cue,
and therefore will not cause a lever release.
The circuit then naturally
uses the conservative
cue-responding strategy.
Switching between these two strategies
depending on the outcome
of the previous trial
allows for adaptive control of action timing.

\section{Emulating aging effects}

The spiking neural implementation
of the double integrator model can be used
to emulate the effects of several pathologies.
As a proof of concept,
we emulate the effects of NMDA receptor degradation
in aging \citep{Magnusson1998}.

Feedforward connections
in the adaptive control model (Figure~\ref{fig4})
use a time constant of 10 ms,
which is consistent with the time constant
of the postsynaptic current
induced by uptake of neurotransmitter
by AMPA receptors \citep{Spruston1995}.
Recurrent connections use a time constant of 50 ms,
which is consistent with the PSC
of NMDA receptors \citep{Sah1990}.
To emulate the effects of aging on NMDA receptors,
we decrease the synaptic strengths
on these recurrent connections.

We hypothesize that this will have the same effect
as adding a small damping term, $D < 0$,
in equation \eqref{dint-full}, resulting in
\begin{align} \label{dint-damped}
  \dot{x}_1 &= u + E + O - R x_1 - D x_1 \nonumber \\
  \dot{x}_2 &= \alpha x_1 - R x_2 - D x_2.
\end{align}
This hypothesis can be tested directly
in the double integrator model,
and its effects on behavior tested
in the adaptive control circuit.

\section{Theoretical results}

\subsection{Model dynamics during the simple RT task}

The dynamical system described by equation \eqref{dint-full}
exhibits the desired behavior during simulation
of the simple RT task.
Figure~\ref{fig5} provides a breakdown of
the trajectory and vector field
throughout a correct, premature, and error trial.
Note that the oscillation is omitted
in this figure.
The final position of
the trajectory becomes the starting position
of the next trial, and therefore
provides information about
the outcome of the last trial.
It is clear from these trajectories
that the time of the cue
can be predicted,
as the ACC is only in the predictive state
when the cue occurs on correct trials.
It is also clear that the
outcome of the previous trials
is easily decoded based
on the state after the previous trial.
Note that premature trials
and late trials cannot be distinguished
on the next trial;
this is consistent with the results of
\citet{Narayanan2009}.

\subsection{The aged model loses adaptivity}

Figure~\ref{fig6} shows trajectories for
correct, premature, and late trials
for the aged model (i.e., the model with a damping term,
equation \eqref{dint-damped}).
As the damping term increases, the final position
of the trajectories on error trials
becomes closer to the final position of correct trials,
indicating that increased damping
makes distinguishing between correct and error trials
progressive more difficult.
Additionally, the distance (in state space) between
the state at which the cue occurs
and other states visited in the trajectory
decreases with increased damping,
which results in more difficulty
predicting the time of the cue
following correct trials as well.
Timing the foreperiod is therefore
a poor strategy with significant aging
(i.e., with $D \le 0.05$).

\section{Simulation results}

\subsection{Spiking networks follow predicted trajectories}

Figure~\ref{fig7} shows that
the decoded values of a spiking neural network
with the same dynamics as equation \eqref{dint-full}
also take a similar trajectory through
state space in correct, premature, and late trials.
Despite the introduction of the oscillation
and the additional noise present
in spiking neural systems,
behaviorally relevant states
are significantly separated
in the decoded state space of the spiking model.
Therefore, the cue can be accurately predicted,
and the outcome of the previous trial
has a significant effect
on the trajectory taken during a trial.

\subsection{Decreasing recurrent weights is analogous to damping}

Figure~\ref{fig8} shows that
the decoded values of a spiking neural network
in which recurrent connections are degraded
by a small amount
take similar trajectories through state space
as the damped model (Figure~\ref{fig6}).
This supports the hypothesis
that a small damping term
has a similar effect as
the degraded NMDA receptors seen in aging
in the double integrator model.

\subsection{Double integrator model has the same principal components as experimental data}

In order to perform an identical
principal component analysis
on the simulated data as was done
on the experimental data,
we randomly sampled 174 neurons
with firing rates $> 1$ Hz from
the spiking double integrator model.
The spike trains of these neurons
were convolved with a Gaussian filter ($\sigma = $ 25 ms)
and then normalized to Z-scores.
The results are shown in Figure~\ref{fig9}.
The simulated principal components
are very similar to the experimental principal components;
$R^2 = 0.96$ and $R^2 = 0.895$ for
the first two components in the post-correct case,
$R^2 = 0.979$ and $R^2 = 0.877$ for
the first two components in the post-premature case,
and $R^2 = 0.967$ and $R^2 = 0.862$ for
the first two components in the post-late case.

\subsection{Adaptive control model has similar performance as experimental subjects}

The parameters of the adaptive leaky integrate-and-fire neurons
used in all models are randomly assigned
from biologically determined distributions.
While the differences in neural parameters
has little effect on the principal components
of the double integrator model,
they can have a significant effect
on performance and reaction times
in the simple RT task.
Similarly, individual experimental subjects
have varying performance levels and reaction times
in the task.
For these reasons, we analyze
each experimental subject and instance of the model separately.

Figure~\ref{fig10} shows the performance
of the experimental subjects,
the model using only the cue-responding strategy,
and the full adaptive model.
The length of the simulations were
varied to approximately match the number of trials
as the experimental subjects.
The mean of the median reaction times
is 272$\pm$48ms for the experimental subjects,
368$\pm$34ms for the cue-responding models,
and 260$\pm$100ms for the adaptive control model.
The mean performance (\% correct trials) is
70.8\% for experimental subjects,
96.9\% for cue-responding models,
and 84.0\% for adaptive control models.
Therefore, the adaptive control model
has similar reaction times as the experimental subjects.
The cue-responding model is significantly slower,
but makes fewer errors than the adaptive control model.

\subsection{Aged model is slower and less variable}

When the aged double integrator model
is used in the adaptive control network,
the mean performance is 89.5\%,
and the mean median reaction time
is 375$\pm$47ms (see Figure~\ref{fig10} for individual results).
These results are closer
to the results of the cue-responding strategy,
suggesting that the model emulating
some effects of aging
uses the conservative cue-responding strategy
over the time prediction strategy
more often than the non-aged adaptive control network.

The results indicate that the double integrator model
is allowing for adaptive control of action timing.
The ability to predict the time of the cue
produces faster reaction times at the cost
of additional premature errors.
The model that enables this adaptive control
has very similar neural dynamics
as the experimental ACC.

One important difference between
the experimentally recorded and simulated ACC
is in the amount of variance
explained by the principal components
and the lack of noise in the
spike density function (see Figure~\ref{fig9}).
The relative cleanness of the simulated data
is expected for several reasons.
Primarily, we expect that the ACC
is sensitive to many other factors
than are represented in the simulation.
The simulated data is more analogous
to the 174 neurons in the entire ACC
that are modulated the most during the simple RT task,
rather than the 174 neurons that happened
to be recorded during the experiments.
This could be simulated
by either injecting
noise into the simulation,
or by representing additional
randomly varying dimensions.
Additionally, the adaptive
leaky integrate-and-fire neuron model used in this study
has a more regular spiking pattern than
biological neurons,
so using a more detailed neural model
would provide a closer match
between experiment and simulation.

An interesting feature of the model
is that there is significant randomness
involved in model generation,
and no learning occurs once the model is generated.
While each instance of the model
can be thought of an individual
from the population of subjects,
one can also think of each model instance as representing
a snapshot during a subject's
learning of the task.
Therefore, the poor performance of, for example,
adaptive control model instance 10 (see Figure~\ref{fig10})
can be explained at least two ways.
It may be a subject
whose recurrent connections are too strong,
and therefore is a poor performer in the task,
or it may be a subject
who has not yet learned how to properly time
the interval, and after more trials would,
with learning, be able to make less premature errors.

The results from the aged model
form an important prediction of
how aged subjects will perform in the simple RT task.
It is notable, however, that the aged model
does not make additional premature errors;
in fact, premature errors are completely eliminated.
This is due to the targeted manner in which
aging is emulated (i.e., NMDA receptor degradation).
If the model is more generally damaged,
for example to emulate an ACC inactivation study,
we would predict that the effects
would be different than in the aging example.
In the case of overall ACC damage,
we hypothesize that the effect is more similar
to decreasing the ``resolution'' of the state space.
That is, it becomes harder to differentiate
similar states, though the system
still takes the same approximate trajectory
through the state space.
In this case, it becomes difficult
to predict the time of the cue,
and therefore additional premature errors would occur.

While we have tested our mathematical model
of ACC function in the context of
control of action timing in a simple RT task,
the tracking of behavioral variables
in a dynamical system could be used
to perform many of the other high-level functions
that the ACC is theorized to perform.
For example, in the commonly studied Stroop task,
the state space being tracked by
the ACC could be the difference between
the response suggested by word-reading and the
response suggested by the color;
the context that colors should be read
would be reflected in the passive dynamics
of the ACC, which would push the ACC
to increasingly drift towards a state
that compensates for the word-reading response.

In conclusion, we have shown that neural integration
is a key mechanism in the ACC's
ability to control the timing of actions.
Analysis of the activity of a neural integrator
model closely matches that of the experimentally recorded ACC.
That model is sufficient to allow
a simple cue-responding network
to react more quickly at the cost
of additional errors,
which matches the performance of experimental subjects.
When the model is manipulated
in a similar way as the aged brain degrades,
performance degrades and forms a prediction
of how aged experimental subjects may perform.

\bibliographystyle{plainnat}
\bibliography{jneurosci2013}

\section*{Supplementary material}

@@

The simple reaction time task
(section \ref{sec:simple-rt}; Figure~\ref{fig1})
is implemented both experimentally
and in a simulated environment.

Experimentally, Long-Evans rats perform the task
in a MedPC box containing a lever,
house lights, and a spout for delivering water reward.
Rats were trained by first
associating reward with lever presses,
and then introducing progressively longer
delay periods until they reached the goal delay
of 1 second.
They were considered to have learned the task
when they performed the task correctly $>60\%$
of the time.

Theoretically, this study was performed
in the Nengo simulation environment.\footnote{Nengo is open source
  software available at \url{http://www.nengo.ca}.}
The environment was modeled using a
a finite state machine to track task state,
and scalar values between -1 and 1
to indicate the state of
house lights, lever position, auditory cue, and reward.

\subsection{Experimental data}

The experimental data analyzed in this study
is the same data presented in
\citet{Narayanan2009}.
The data consists of 174 neurons
recorded from anterior cingulate cortex with microwire arrays
in twelve male Long-Evans rats.
For more details on the data collection,
refer to the original paper.

\subsection{Simulation details}

All simulations were run
in the Nengo simulation environment
through Python scripts.
Those scripts are available
at \url{https://github.com/tbekolay}.

\subsubsection{Neural Engineering Framework} \label{sec:nef}

Nengo simplifies the construction
of spiking neural networks
using the Neural Engineering Framework (NEF; \citealp{Eliasmith2003}).
The NEF makes the assumption
that populations of spiking neurons
represent real-valued vectors,
and use a least-squares optimal method
to transform those representations
through the connections between
populations of neurons.
A population can also be recurrently connected
in order to produce dynamics,
such as the neural integrators
employed in this study.

Briefly, the representation scheme
is similar to population coding,
as proposed by \citet{Georgopoulos1986}
but extended to $n$-dimensional vector spaces.
Each neuron in a population is sensitive
to a particular direction,
called the neuron's \textit{encoder}.
The activity of a neuron can be expressed as
\begin{equation}
	a = G[\alpha \mathbf{e} \cdot \mathbf{x} + J_{bias}],
\end{equation}
where $G[\cdot]$ is the nonlinear neural activation function,
$\alpha$ is a scaling factor (gain) associated with the neuron,
$\mathbf{e}$ is the neuron's encoder,
$\mathbf{x}$ is the vector to be encoded, and
$J_{bias}$ is the background current of the cell
when $\mathbf{x} = 0$.
The currently encoded value, $\mathbf{\hat{x}}$,
can be estimated linearly.
\begin{equation}
  \mathbf{\hat{x}}(t) = \sum_i \mathbf{d}_i a_i(t),
\end{equation}
where $\mathbf{d}_i$ is the decoder,
and $a_i$ is activity of neuron $i$.

Neural activity is interpreted as a
filtered spike train.
\begin{equation}
  a_i(t) = \sum_s h(t - t_s) = \sum_s e^{-(t - t_s) / \tau_{PSC}},
\end{equation}
where $h(\cdot)$ is the exponential filter
applied to each spike,
and $s$ is the set of all spikes occurring
before the current time $t$.

The decoders are found through a least-squares minimization
of the difference between the decoded estimate
and the actual encoded vector.
\begin{equation}
  \mathbf{d} = \Upsilon^{-1} \Gamma \hspace{1.8em}
  \Gamma_{ij} = \int a_i a_j dx \hspace{1.8em}
  \Upsilon_j = \int a_j \mathbf{x} dx.
\end{equation}

Connection weights computing a function
between two populations connected in a feedforward manner
$f(\mathbf{x})$ can be determined by
solving for a set of decoding weights
for that function,
\begin{equation}
  \mathbf{d}^{f(\mathbf{x})} = \Upsilon^{-1} \Gamma \hspace{1.8em}
  \Gamma_{ij} = \int a_i a_j dx \hspace{1.8em}
  \Upsilon_j = \int a_j f(\mathbf{x}) dx,
\end{equation}
and then computing the following weight matrix, $\omega$;
\begin{equation}
	\omega_{ij} = \alpha_j \mathbf{e}_j L \mathbf{d}^{f(\mathbf{x})}_i,
\end{equation}
where $i$ indexes the input population,
$j$ indexes the output population,
and $L$ is a linear operator.

Dynamical systems of the form
$\dot{\mathbf{x}} = A(\mathbf{x}) + B(\mathbf{u})$
can be created by connecting a population
to itself with a weight matrix
that computes the function $f(\mathbf{x}) = \mathbf{x}$,
setting $L = A$ on the recurrent connection,
and receiving input $\mathbf{u}$
from other neural populations
and setting $L = B$ on those connections.

@@

In addition to the description in the main text,
it should be noted that \citeauthor{Singh2006}
discuss four possible
realizations of their double integrator model,
all of which produce the same dynamics
in the final output population.
The realizations differ in whether
the signals are internally or externally generated,
and in whether a single multidimensional integrator
or multiple one-dimensional integrators are used.

For ease of implementation,
we used multiple one-dimensional integrators
because the resulting network is more stable than
a network with a single multidimensional integrator
\citep{Singh2006}.
No microstimulation was done
during the behavioral experiments
so it is difficult to determine
if the experimental ACC
can also be thought of as
a system of coupled integrators;
preliminary simulations using
a single multidimensional integrator
produced PCs that matched no better
than the coupled integrator model,
and often worse.
